String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-29 2023-1-19/sections/overview.tex
overview finished

Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters. Then we apply the determinization and minimization algorithm of NFAs to $\cB$, resulting into an automaton $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.5\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cC$ after ignoring the characters in $\aut_1 \times \aut_2$ and applying determinization and minimization}
  \label{fig:overview:product:reduced}
\end{figure}

\zhilin{stopped here}

\hide{
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
}

@@

Sometimes the string constraints are satisfiable, and the strings in the solution have a short length. Such a solution may be quickly explored by derivation rules in DPLL(T)-based string solvers but slowly explored by our approach. To improve efficiency on satisfiable string constraints, we propose a light-way heuristic that tries to find a solution in the under-approximation of the string constraints. The main idea is to explore paths within an exact length and manually compute the registers' values, rather than reduce CEFA to heavy linear integer arithmetic. For example, consider the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c$, we can explore paths within length 10 and get a satisfiable solution $x = cccccccccc$. 
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example_simplified.tex}
  \caption{The simplified CEFA handling the difficult string constraints}
  \label{fig:overview:simplified}
\end{figure}

>>>>

Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters. Then we apply the determinization and minimization algorithm of NFAs to $\cB$, resulting into an automaton $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}. 
Note that $\cC$ contains only three states $q'_0, q'_1, q'_2$ and six transitions $\tau'_1, \cdots, \tau'_6$, thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA-reduced}
\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.3\textwidth]{sections/overview-cefa-reduced.pdf}
  \caption{Reduced automaton $\cC$}
  \label{fig:overview:product:reduced}
\end{figure}
Furthermore, we can reduce the number of registers by merging several registers into one, if they are updated with the same value in every transition of $\aut_1 \times \aut_2$.

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from CEFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially we set $B$ to $1$. If there does not exist a $B$-bounded satisfiable assignment, then we let $B:=B+1$ and continue the search, until finding a satisfiable assignment or reaching a given time limit. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $1$ is already sufficient to find a solution for $\aut_1 \times \aut_2$ (with $r_0 > 120$ ignored).

%%%%%%%%%%%%%%%%%original texts by Denghang%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%original texts by Denghang%%%%%%%%%%%%%%%
\hide{
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}

@@

Sometimes the string constraints are satisfiable, and the strings in the solution have a short length. Such a solution may be quickly explored by derivation rules in DPLL(T)-based string solvers but slowly explored by our approach. To improve efficiency on satisfiable string constraints, we propose a light-way heuristic that tries to find a solution in the under-approximation of the string constraints. The main idea is to explore paths within an exact length and manually compute the registers' values, rather than reduce CEFA to heavy linear integer arithmetic. For example, consider the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c$, we can explore paths within length 10 and get a satisfiable solution $x = cccccccccc$. 
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example.tex}
  \caption{The CEFA handling the difficult string constraints}
  \label{fig:overview:orgin}
\end{figure}
\begin{figure}[ht]
  \centering
  \import{figures}{overview_example_simplified.tex}
  \caption{The simplified CEFA handling the difficult string constraints}
  \label{fig:overview:simplified}
\end{figure}
}
%%%%%%%%%%%%%%%%%original texts by Denghang%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%original texts by Denghang%%%%%%%%%%%%%%%
